Death by Dropdown?<br>Engineer Insightful Shiny Apps<br>with Behavioral Science

### 2025-09-15 conference-materials/2025_posit-conf/draft.qmd
Update presentation notes with improved timing and clearer talking points

---
title: "Death by Dropdown?<br>Engineer Insightful Shiny Apps<br>with Behavioral Science"
author: "Jeremy Winget, PhD"
date: "2025-09-07"
date-format: "MMMM D, YYYY"
execute:
  echo: true
  eval: true
format:
  revealjs:
    code-block-background: true
    highlight-style: github
    slide-number: false
    transition: fade
    auto-animate: true
    width: 1600
    height: 900
    footer: "posit::conf(2025) | jrwinget\\.com"
    title-slide-attributes:
      data-background-image: "img/posit-conf.png"
      data-background-position: "top left"
      data-background-repeat: "no-repeat"
      data-background-color: "#213d4f"
      data-background-opacity: "0.25"
      data-background-size: "45%"
---

@@

::: {.notes}
- 10 seconds
- Ever open a dashboard with 18 filters across 6 separate tabs...and try to figure out where to even start?
- I mean, that's a lot of data and different ways slicing it to get a full understanding of what's going on...so what do you do?
- That is...if you haven't made your career out of exploring data sets like most executives making decisions...what do you do?
- [CLICK] Maybe you feel a bit like this...
- [PAUSE]
- [CLICK] Or maybe you have some bigger feelings about it...
- I get the feeling I'm not the only one who's been here before
:::

@@

*Contrary to popular belief, most psychologists work with data. Not clients.*
:::

::: {.column width="50%" style="margin-top: var(--space-8);" .fragment .center}
### My Journey {style="color: var(--indigo-light) !important;"}

**Applied Social Psychology** → <br>
**Full Stack Engineer** → <br>
**Led Shiny teams at scale** → <br>
**Created {bidux} package**
:::
::::

::: {.notes}
- 45 seconds
- Alright, let's get honest with each other for a moment. Quick show of hands - who here has created a user interface with more than 10 filters? Don't be shy!
- How about this one - who's ever had a client or user ask for "just one more dropdown" when you knew it would make things worse?
- Or here's my personal favorite - who's built a Shiny app with more selectInput() functions than actual insights?
- [CLICK] If any of these sound familiar, then welcome to what I like to call my dashboard support group! You're among friends here.
- Seriously, think of our time together today as a little group therapy session for dropdown overload. We've all been there, and there's no judgment in this room.
- [PAUSE] Let me introduce myself properly - Hi, I'm Jeremy! I'm a psychologist by training, but probably not the kind you're thinking of.
- Here's something that surprises people - contrary to popular belief, most psychologists work with data, not individual clients. We use it to study behavior patterns, understand how people make decisions, and figure out why humans do the puzzling things we do.
- I have a PhD in Applied Psychology where I specifically studied how people and groups process information and make decisions under pressure - which, it turns out, is exactly what happens when someone opens your dashboard!
- I transitioned from academia into industry where I got involved in leading Shiny teams at scale, and I've spent nearly 10 years building enterprise apps and dashboards that people actually use and find valuable.
- But I have to be honest - it wasn't always that way. I've built MANY dashboards that flopped spectacularly, even though they were technically perfect. Every single one of those problems I mentioned? Yeah, I've made them all.
- And that's exactly what led me to my core realization about why dashboards succeed or fail...
:::

@@

> Most dashboard failures aren't technical...
>
> They're psychological.

::: {.notes}
- 15 seconds
- Most dashboard failures aren't technical, they're psychological
- That might sound obvious, but it's a critical distinction...and one that we as developers can also easily address in our workflows
- Let me show you what I mean...and how technically great products can still fail spectacularly
:::

@@

::: {.text-box-overlay-dark .absolute left="-200" bottom="100" .fragment}
### Real client. Real dashboard. Real problems.

@@

::: {.notes}
- 45 seconds
- I worked with a client who had this 'comprehensive' dashboard
- Blinded here for confidentiality...but incredibly close to the actual design
- 18 filters, 6 tabs, every possible view, nearly zero adoption
- Why this matters: This represents thousands of developer hours and client budget wasted on technically perfect but psychologically broken software
- The client was frustrated bc they had spent months building this thing, but no one used it
- Users were initially curious, but quickly overwhelmed...and eventually, they just vanished
- As developers, we can relate...we've all built features nobody uses
- But here's why this matters more than frustration:
    1. Technical debt from unused complexity compounds over time
    2. User abandonment patterns replicate across similar projects
    3. Without data-driven insights, we keep making the same UX mistakes
- [CLICK] So we ran telemetry for 30 days...
- This data changed everything about how we approached the dashboard design
:::

@@

::: {.notes}
- 45 seconds
- Look at this journey of despair
- [update notes here]
- This is behavioral science **screaming** at us that something is wrong here
:::

@@

::: {.notes}
- 1 minute
- Behavioral science studies how people think, feel, and act...often in the context of decision making
- Surprise...we're not rational...but we are still predictable and fairly consistent when we pay attention to the right drivers
- Let me show you a classic example that mirrors our dashboard problem...Sheena Iyengar & Mark Lepper's famous jam study at Columbia
- They studied people in the field at a fancy grocery store...think Whole Foods or Trader Joe's
- These shoppers walked by a display stand with either 24 jams...(extensive choice condition)...or 6 jams (limited choice condition)
- More choice drew people in...60% stopped at the big display vs. 40% at the small one
- But fewer choices drove action...nearly 30% bought jam after the 6-jam booth, compared to only 3% after the 24-jam booth
- More choice attracted browsers...but paralyzed buyers
- So what do we learn from this?
    - Too many options look appealing but kill follow-through...exactly what happens with dashboards overloaded with filters
:::

@@

::: {.notes}
- 30 seconds
- Now the paradox of choice is just one of many behavioral science concepts that directly apply to dashboards
- [CLICK] Let me show you specific examples of how these apply
- Cognitive Load: Instead of 18 filters overwhelming users, start with 3 essential ones and hide advanced options
- Progressive Disclosure: Show high-level KPIs first, then let users drill into details, then comparisons
- Framing Effects: The same revenue number can be "up 15%" or "missed target by 5%" - completely different user reactions
- Choice Architecture: Smart defaults, logical grouping, visual hierarchy guides users naturally
- These aren't just theories...they're actionable engineering principles
:::

@@

![](img/bid-framework.png){style="box-shadow: none !important; margin-left: 10%"}
:::
::::

::: {.notes}
- 45 seconds
- Think of building dashboards like planning the perfect road trip with friends. You know that feeling when you're mapping out an adventure - equal parts excitement and careful preparation?
- And speaking of adventures - for those Grateful Dead fans in the audience, building user-friendly dashboards really can be a "long, strange trip." But unlike Jerry Garcia's journeys, ours actually have a clear destination and users who need to get there!
- Just like any road trip, you need to interpret where you're going, notice those warning lights on the dashboard, anticipate weather changes...
- [CLICK] This is what I call the BID framework - your roadmap for dashboard success
- Five iterative stages that prevent those dashboard disasters we've all experienced. Think of it as your GPS for user experience.
- Like any memorable road trip, it's not just about reaching your destination - it's about creating a framework that makes the journey enjoyable and the discoveries along the way meaningful for your users.
:::

@@

::: {.column width="45%" .fragment style="margin-top: 150px;"}
### *Your behavioral scientist in the console* {style="text-align: center !important; color: var(--indigo-light) !important;"}
:::
::::

::: {.notes}
- 30 seconds
- So what exactly is bidux? Think of it as having a friendly behavioral scientist sitting right there in your console, ready to help whenever you're building dashboards.
- It works with any Shiny app you've already built - no need to start over. Whether you have telemetry data or not, it can still provide valuable insights and auto-suggest improvements based on solid behavioral science principles.
- [CLICK] The best part? It translates all those academic psychology concepts into practical, actionable code suggestions that you can implement today. It's like having a psychology PhD on your development team, but one that actually speaks R!
:::

@@

::: {.notes}
- 1 minute
- "Start with why - what question does this dashboard answer?"
- [CLICK] First highlight: central_question - this drives everything else
- [CLICK] Second highlight: data_story - this creates narrative structure
- [CLICK] Third highlight: user_personas - this grounds design in real needs
- Why this matters for your daily work: Clear purpose drives every technical decision you make
- Most devs think they do this, but telemetry says otherwise. We jump straight to technical implementation
- The cost: Every filter you add without purpose increases cognitive load exponentially
- This isn't just UX theory - it's engineering efficiency. When you clarify purpose first:
    1. You write less code because you know what to exclude
    2. Your data queries are more focused and performant
    3. User testing becomes objective rather than subjective
    4. You can defend design decisions with evidence, not opinions
- `bid_interpret` creates a structured object that feeds into later stages, making behavioral science actionable in your R workflow
- Think of it as writing better requirements that actually connect to user psychology
:::

@@

::: {.notes}
- 1 minute
- Notice finds the concepts behind the friction points
- Why this stage matters for developers...It turns user complaints into actionable engineering insights
- Look at this heatmap...14 filters basically going unused...4 are completely ignored
- This isn't just usage data...it's technical debt visualization
- What this means for your codebase:
    1. Those unused filters are maintenance overhead with zero ROI
    2. Database queries could be running for features nobody touches
    3. UI complexity is slowing rendering times without value
    4. Future feature requests will compound these problems over time
- `bid_notice` auto-suggests behavioral theories to explain patterns...or power users can override them with specific concepts
- This transforms "users don't like it" into "choice overload is killing task completion"
- Now you have a specific problem to solve with code
:::

@@

::: {.notes}
- 1 minute
- "Anticipate cognitive biases"
- "Same data tells different stories"
- "Marketing sees progress, Finance sees gap"
- "bidux auto-generates these guards"
- "Framing changes everything"
:::

@@

::: {.notes}
- 1.5 minutes
- "Structure provides implementation roadmap"
- "Specific code recommendations"
- "Progressive disclosure reduces cognitive load"
- [CLICK] "Follow natural eye patterns"
:::

@@

:::: {.r-stack}
::: {.fragment .fade-in}
```{r}
validate_stage <- bid_validate(previous_stage = structure_stage)
```
:::

::: {.fragment .fade-up style="width: 70%"}
![](img/ui-summary.png)
:::
::::

::: {.notes}
- 1 minute
- "Validate ensures understanding"
- "Plain language summaries"
- "Auto-generates checklist"
:::

@@

![](img/ui-before2.png)

::: {.notes}
- 45 seconds
- [update notes here]
- "Δ (pp)"" is absolute percentage-point change
- Relative change is After/Before - 1, then rounded
- [CLICK] More users. Deeper engagement. Stronger retention.
:::

@@

::: {.notes}
- 45 seconds
- [update notes here]
- "Δ (pp)"" is absolute percentage-point change
- Relative change is After/Before - 1, then rounded
- [CLICK] More users. Deeper engagement. Stronger retention.
:::

@@

::: {.notes}
- 30 seconds
- "Real results from 12 enterprise deployments"
- Why these metrics matter for your engineering career:
    1. 82% reduction in time to insight = users find value faster = better adoption metrics for your projects
    2. 143% improvement in task completion = fewer support tickets = less maintenance overhead
    3. These improvements compound: happy users request thoughtful features, not band-aid fixes
- "This works across industries" - the psychological principles are universal, but implementation adapts to your specific R/Shiny tech stack
- Bottom line: Better user psychology = better engineering outcomes = more impactful projects in your portfolio
:::

@@

::: {.notes}
- 1 minute
- "Complete workflow in one pipe - works with or without telemetry"
- [CLICK] With Telemetry tab first:
- [CLICK] First highlight: Load and analyze existing data
- [CLICK] Second highlight: The 5-stage BID pipeline
- [CLICK] Third highlight: Generate report
- [CLICK] Without Telemetry tab:
- [CLICK] First highlight: Manual problem definition
- [CLICK] Second highlight: Same 5-stage pipeline
- [CLICK] Third highlight: Text output for tickets/PRs
- bid_telemetry() provides a tidy tibble of issues compatible with dplyr
- bid_notice_issue() is the bridge from a single issue to a BID Notice stage
- The 5-stage chain culminates in bid_report(), which can produce text, HTML, or markdown
- Output is a concise report you can paste into PRs or tickets
- "Everything you need to fix your dashboard"
:::

@@

::: {.notes}
- 1 minute
- "Complete workflow in one pipe - works with or without telemetry"
- [CLICK] With Telemetry tab first:
- [CLICK] First highlight: Load and analyze existing data
- [CLICK] Second highlight: The 5-stage BID pipeline
- [CLICK] Third highlight: Generate report
- [CLICK] Without Telemetry tab:
- [CLICK] First highlight: Manual problem definition
- [CLICK] Second highlight: Same 5-stage pipeline
- [CLICK] Third highlight: Text output for tickets/PRs
- bid_telemetry() provides a tidy tibble of issues compatible with dplyr
- bid_notice_issue() is the bridge from a single issue to a BID Notice stage
- The 5-stage chain culminates in bid_report(), which can produce text, HTML, or markdown
- Output is a concise report you can paste into PRs or tickets
- "Everything you need to fix your dashboard"
:::

@@

::: {.notes}
- 30 seconds
- Here's the exciting part - you can get started with bidux right now! It's available on CRAN, so installation is as simple as any other R package.
- The beautiful thing is it works immediately, even without telemetry data. You can start getting insights and suggestions for your existing dashboards today.
- [CLICK] I've also prepared plenty of resources to support your journey. These slides will be available in the BID framework repository, and I'll be posting them on my website as well.
- I genuinely believe the R and Shiny community is one of the most supportive and collaborative communities out there, so please join us! Feel free to connect with me on Bluesky, LinkedIn, or email - I'd love to hear about your experiences and learn from your dashboard challenges too.
:::

@@

::: {.notes}
- 20 seconds
- If you take away just one thing from our time together today, let it be this: Dashboards don't need more features - they need fewer obstacles.
- Every dropdown, every filter, every choice we add should bring users closer to insight, not further from it.
- [CLICK] And here's the thing - we can absolutely fix this together. The tools exist, the knowledge exists, and most importantly, you all clearly care about creating better experiences for your users.
- [CLICK] Before I close, I want to offer my heartfelt thanks to the posit::conf organizers for creating this incredible space for learning, to the broader Shiny community for pushing the boundaries of what's possible with data visualization, and most importantly, to all of you for caring enough about your users to be here learning how to serve them better.
- Now, I'd love to hear your questions and learn about the dashboard challenges you're facing!
:::

>>>>

---
title: "Death by Dropdown?<br>Engineer Insightful Shiny Apps<br>with Behavioral Science"
author: "Jeremy Winget, PhD"
date: "2025-09-17"
date-format: "MMMM D, YYYY"
execute:
  echo: true
  eval: true
format:
  revealjs:
    code-block-background: true
    highlight-style: github
    slide-number: false
    transition: fade
    auto-animate: true
    width: 1600
    height: 900
    footer: "posit::conf(2025) | jrwinget\\.com"
    title-slide-attributes:
      data-background-image: "img/posit-conf.png"
      data-background-position: "top left"
      data-background-repeat: "no-repeat"
      data-background-color: "#213d4f"
      data-background-opacity: "0.25"
      data-background-size: "45%"
      data-notes: |
        - [10 seconds; smile warmly at audience]
        - Good morning everyone! Let's jump right in...
---

@@

::: {.notes}
- [30 seconds]
- Who here has opened a dashboard with 18 filters across 6 tabs...and wondered where to even start?...Or maybe what they were even looking at?
- I see some knowing looks
- Now I'd like you to imagine you're NOT a data scientist...
- You're an executive who needs to make a decision in the next 10 mins
- [CLICK] This might be you...frantically clicking, hoping something makes sense
- [CLICK] Or maybe you have some...bigger feelings
- The truth is...we've all probably felt this way towards a dashboard...and it's never a good feeling
- [end by 40 secs]
:::

@@

*The public sees therapists, but the discipline is powered by data.*
:::

::: {.column width="50%" style="margin-top: var(--space-8);" .fragment .center}
### My Journey {style="color: #8D9CEF !important;"}

**Applied Social Psychology** → <br>
**Full Stack Engineer** → <br>
**Led Shiny teams at scale** → <br>
**Researcher / OSS Developer**
:::
::::

::: {.notes}
- [60 seconds]
- So...welcome to my dashboard support group!...Let's do a quick check-in
- Raise your hand if you've designed a UI with 10+ filters
    - Keep it up if you've gone over 20?
- How about this...who's had a user ask for "just one more dropdown" when you KNEW it was just too much?
- [CLICK] If you're nodding along...you're among friends...No judgment here
- [CLICK] I'm Jeremy...and I've been there too
- I'm a psychologist by training...but probably not the kind you're thinking of
- See...a lot of folks think psychology is all about couches and therapy...but the discipline is powered by data
- [CLICK] My journey took me from studying how groups make decisions under pressure...to building the systems that caused that pressure...to finally figuring out how to fix them
- And after years of building dashboards that were technically great...but didn't seem to land with users...I had this realization
- [end by 1:40 mins]
:::

@@

> A lot of dashboard failures aren't technical...
>
> They're psychological.

::: {.notes}
- 20 seconds
- A lot of dashboard failures aren't technical...They're psychological
- They don't fail bc of the wrong chart type or database
- They fail because humans need to use them...and humans are complicated
- But we can engineer for psychology just as precisely as we engineer for performance
- Let me show you what I mean
- [end by 2:00 mins]
:::

@@

::: {.text-box-overlay-dark .absolute left="-200" bottom="100"}
### Real client. Real dashboard. Real problems.

@@

::: {.notes}
- [45 seconds]
- This is a dashboard from an actual client...details blinded of course, but still accurate
- They had built a really impressive dashboard...
- Lots of filters...customized views...and some really cool visualizations for deep dives
- The client loved this bc..."users can analyze anything they want!"
- Except...they never did
- The opened the app...but didn't stay long before they completely deserted it
- They were finding creative workarounds...like Excel
- [CLICK] So we decided to roll out a minor update and collect anonymous usage data with telemetry for 30 days to get a better idea of what users were actually doing
- [end by 2:45]
:::

@@

::: {.notes}
- [35 seconds]
- And in this case...it painted a user journey like this cliff of despair
- 95% of users visited the URL...fantastic! Shows people are interested
- 88% opened the dashboard...also good...they're getting past the landing page
- But here's where it gets brutal...only 52% applied even one filter. Just one!
- And only 20% returned within 7 days
- This chart is behavioral science screaming "STOP DOING THIS!"
- But what needs fixing...and how do we do it?
- [end by 3:20]
:::

@@

::: {.notes}
- [75 seconds]
- With behavioral science!...which is the study of how humans think, feel and act...often in the context of decision making
- The field itself is a blend of psychology, economics, neuroscience, and others...
- But across all those areas we find people are often irrational...but in consistent ways
- For us as developers...this is actually great news!
- It means user behavior is predictable...and we can design for it
- [CLICK] Let me share a study that demonstrates our death by dropdown problem perfectly
- [CLICK] In this study...two researchers set up a sampling booth at a grocery store...and they varied the number of jams they on display
- [CLICK] In the big display condition...there was were 24 jams on the table...and in the small display condition...there were 6 jams
- [CLICK] The big display with 24 jams attracted more shoppers than the small display with 6 jams
- BUT...only 3% of folks who stopped at the larger display bought jam...whereas 30% of people bought jam in the small display condition
- [CLICK] That's a 10x conversion difference!
- So more choice attracted attention but paralyzed action...and filter-heavy dashboards do the exact same thing
- [end by 4:30]
:::

@@

::: {.notes}
- [75 seconds]
- Now...the paradox of choice is just one of many behavioral science concepts we can apply to dashboards
- But these aren't abstract theories...they're actually engineering principles we can leverage
- Cognitive Load...this means too many choices overwhelm users' mental capacity
    - Like RAM...your users' brains have processing limits...every filter costs processing power
- Progressive Disclosure...this means revealing complexity gradually as users need it
    - Show vital information first...and reveal more as needed
    - Like lazy loading for human cognition
- Framing Effects...this is how the same data can tell completely different stories
    - "Revenue up 15%" versus "Missed target by 5%"
    - Identical data...but the Proj Manager might see success while the CFO sees failure
- Choice Architecture...this means guiding the user without restricting them
    - Hide rarely used filters, use smart defaults
    - Like good API design...guide users down the happy path
- [CLICK] In practice...this means starting with 3 filters not 18...showing summaries before details...and using smart defaults
- This is how we engineer the user experience as carefully as we would the backend
- But what if you don't know behavioral science?
- [end by 5:35]
:::

@@

![](img/bid-framework.png){style="width: 80%; box-shadow: none !important; margin-left: 20%"}
:::
::::

::: {.notes}
- [45 seconds]
- Well...I bet you do know something about road trips
- And engineering apps with behavioral science is a lot like planning a road trip
- You need to know your destination...anticipate the hazards...and make sure everyone gets there in one piece
- That's where the Behavioral Insight Design (BID) framework comes in...[CLICK]
    - It's like having GPS for your behavioral science journey
- Five simple stages that guide you through the process...and help prevent those dashboard disasters we've all seen
- What makes BID different is it's not some abstract design theory
- It's a concrete...step-by-step framework built on decades of research
- Each stage feeds into the next...creating this continuous feedback loop that keeps improving your dashboard
- Plus...the whole thing is already bundled into an R package that handles the heavy lifting
- [end by 6:20]
:::

@@

::: {.column width="45%" .fragment style="margin-top: 150px;"}
### *Your behavioral scientist in the console* {style="text-align: center !important; color: #8D9CEF !important;"}
:::
::::

::: {.notes}
- [45 seconds]
- Meet bidux...designed specifically for our workflows as Shiny developers
- It works with any dashboard you've already built...no refactoring, no starting over
- Got telemetry? Great, it'll analyze actual user behavior and give you data-driven insights
- Or...if you don't have telemetry data...you can use simple verbal descriptions during the process
- Whatever inputs you provide...it will use them to auto-suggest improvements based on real research
- [CLICK] Think of it as having a behavioral scientist right in your console...always available, never judges, and doesn't charge by the hour
- Now...let me show you what this process actually looks like
- [end by 7:05]
:::

@@

::: {.notes}
- [65 seconds]
- First...we start with the INTERPRET stage...the why behind the dashboard
- What question does this dashboard answer?
- [CLICK] Because the central question drives the focus for your users
- If you can't state it in one simple sentence...they won't find it in a bunch of dropdowns
- [CLICK] And the data story helps create narrative structure for this
- Humans are hardwired for stories...not pivot tables
- Hook, context, tension, resolution...is the same structure folks are looking for with their key insights...so we build that in from the start
- [CLICK] And finally...the user personas help us devs take the user perspective
- The Product Manager isn't you...they don't think in SQL queries...They think in business outcomes
- So the dashboard needs to bridge that gap
- [CLICK] And when we put it all of that information together...the function creates a summary for you to review before moving on
- If it tells you anything is missing...it could also be a good time to check in with your users
- [end by 8:10]
:::

@@

::: {.notes}
- [75 seconds]
- Bc when we don't consider that context...we end up building critical friction points into our dashboards
- Going back to that opening example...here's what filter usage looked like
- 14 filters basically unused...4 virtually ignored
- This is technical debt visualized
- Each unused filter is maintenance overhead...and there are probably database queries running for nothing
- That's why the NOTICE stage focuses on finding these friction points systematically
- [CLICK] The package auto-suggests relevant behavioral theories with confidence scores
- Here...90% confidence it's Cognitive Load Theory...too many choices paralyzing users
- But it might suggest Choice Overload...Analysis Paralysis...or Decision Fatigue
- Each theory comes with specific, implementable solutions
- This transforms "it's confusing" into specific engineering problems...like "cognitive load exceeds working memory"
- The package also links each theory to specific Shiny components that address it
- So you're not guessing anymore...you're addressing identified psychological barriers with targeted code changes
- [end by 9:25]
:::

@@

::: {.notes}
- [45 seconds]
- [CLICK] The package auto-generates bias mitigations based on your specific dashboard context and user personas
- [CLICK] Look at the specific suggestions...anchoring, framing, confirmation bias
- These aren't abstract concepts...they're bugs in human cognition that affect every user
- [CLICK] Same dashboard, gain framing..."We're making progress!" Optimistic, forward-looking
- [CLICK] Loss framing..."We're behind target!" Creates urgency, drives action
- Same data...but different psychological impact
- Marketing sees growth opportunities...Finance sees risk exposure...both are right, both need different framing
- bidux helps you present both perspectives fairly...or choose the right frame for your audience
- It also suggests things like reference lines for anchoring...confidence intervals to combat overconfidence...and alternative views to fight confirmation bias
- This is proactive debugging for human psychology
- [end by 10:10]
:::

@@

::: {.notes}
- [65 seconds]
- At this point...we know what the dashboard is for, what problems to fix, and what biases to guard against
- Now it's time to STRUCTURE the information flow
- [CLICK] Look at these implementation suggestions...specific bslib and shiny components
- Not just "reduce complexity" but "use bslib::accordion for progressive disclosure" with actual code examples...(with more detailed suggestions coming soon!)
- The package knows which components map to which psychological principles
- [CLICK] See this F-pattern heatmap? This is how users scan dashboards in Western cultures
- Top-left gets 100% attention...bottom-right might as well not exist
- Structure stage puts your most important insights where eyes naturally go
- Critical KPIs or Summaries go on the top...supporting details follow the F-pattern...deep-dive options progressively disclosed
- This is cognitive engineering...you're structuring information to match how the human brain processes visual data
- Your most expensive computations should align with where users actually look
- [end by 11:15]
:::

@@

:::: {.r-stack}
::: {.fragment .fade-in-then-out}
```{r}
validate_stage <- bid_validate(previous_stage = structure_stage)
```
:::

::: {.fragment .fade-up}
![](img/ui-summary.png){.r-stretch}
:::
::::

::: {.notes}
- [40 seconds]
- Finally...we reach the last stage: validate and empower users
- This helps you know the dashboard works as intended...and helps users know what to do next
- It guides you through testing changes and iterating...but also suggests ways to empower user engagement
- For example...executive summaries tailored to personas...detailed wording for analysts, high-level for execs
- This stage generates checklists...checks accessibility...and creates those "aha" moments
- It's the difference between "I looked at the dashboard" and "I know exactly what to do now"
- But Validation isn't the end...it feeds back to Interpret...creating continuous improvement that you can always revisit as needed
- Now let's take a look at the full transformation
- [end by 11:55]
:::

@@

![](img/ui-before2.png){.r-stretch}

::: {.notes}
- [25 seconds]
- Here's our dashboard disaster again...18 filters, 6 tabs, information overload incarnate
- Technically perfect...every edge case handled, every possible view available
- Psychologically broken...users spent more time figuring out the interface than finding insights
- This dashboard had everything except users
- Let's see what behavioral science can do...
- [end by 12:20]
:::

@@

::: {.notes}
- [65 seconds]
- Same functionality...but a transformed user experience
- Clean executive summary at the top...users know immediately what matters and what to do about it
- Progressive disclosure...complexity is available but not required
- Smart defaults based on actual usage patterns...not assumptions
- [CLICK] Now look at these metrics...this is the ROI of behavioral science
- Filter usage up 38%...people are actually exploring data instead of abandoning ship
- Return rate up 75%...they're coming back!
- That's the difference between a proper tool and shelfware
- This isn't magic...or AI...it's engineering for how humans actually think
- We reduced cognitive load...aligned with mental models...and created clear information hierarchy
- [CLICK] More users...deeper engagement...stronger retention...because we removed obstacles instead of adding features
- And as a developer...this also means...
    - Less support tickets because users can figure it out themselves
    - Fewer feature requests because they're finding what they need
    - Better performance reviews because your dashboards driving value
- And that's the difference between building features and products people love
- [end by 13:25]
:::

@@

::: {.notes}
- [45 seconds]
- And here are just some data to back that up
- These aren't cherry-picked results...they the 12 enterprise deployments across industries I've worked on in the last year using the BID framework
- 65 seconds faster to first insight...that's over a minute saved per user, per session
- Multiply that by hundreds of users, daily visits...you're saving hours of collective time
- Task completion up from 48% to 62%...nearly a third more decisions actually getting made
- User satisfaction up 27%...that's the difference between "I have to use this" and "I want to use this"
- Why does this work across industries?...Because human psychology doesn't change across domains
- Your finance users have the same 7±2 working memory slots as healthcare users
- They all suffer from choice overload...
- The framework adapts to your domain...but the principles remain constant
- Better psychology equals better engineering equals happier developers and users
- [end by 14:10]
:::

@@

::: {.notes}
- [40 seconds]
- Here's the really nice part...implementing all this psychology is just a few lines of code
- [CLICK] If you have telemetry...simply load it with bid_telemetry
- [CLICK] Here...I'm filtering to to one of the critical issues identified
- [CLICK] Then adding a simple question the dashboard is trying to answer to contextualize everything that follows
- [CLICK] And then I use bid_notice_issues...a bridge function from telemetry to behavioral insights that auto-populates the following stages
- [CLICK] From there...it's running through the remaining stages...
- [CLICK] And ending with bid_report to optionally generate a markdown report ready for your PR or documentation
- [end by 14:50]
:::

@@

::: {.notes}
- [50 seconds]
- And if you don't have telemetry...that's not a problem either
- [CLICK] Just define your problem manually based on user feedback or your observations
- [CLICK] Create your data story...to structure the dashboard's narrative
- [CLICK] Identify the friction points...even without data...you probably know where users struggle
- [CLICK] Same five-stage pipeline...same actionable insights
- [CLICK] Generate text output perfect for tickets, documentation, or stakeholder communication
- The framework guides you whether you have tons of telemetry or simple user feedback
- It's behavioral science made practical for your daily workflow
- You can run this during development...not just after deployment
- Catch psychological issues before they become user complaints
- This is shift-left for UX...fixing problems before they're problems
- [end by 15:40]
:::

@@

::: {.notes}
- [60 seconds]
- You can literally start using this today
- It's available on CRAN, so installation is as simple as any other R package
- [CLICK] Install & load the library...you know the drill
- [CLICK] Run through the five stages...even with minimal input, you get valuable suggestions
- [CLICK] Use bid_concepts to explore specific behavioral principles...it's like having a psychology textbook on hand for further reading
- Want to understand choice overload? bid_concepts("choice overload") gives you theory, impact, and solutions
- [CLICK] Check the next_steps for actionable improvements...prioritized by impact
- [CLICK] I've created extensive resources...docs, discussions, examples, all designed for the R/Shiny community
- These slides will be available in the BID framework repository...and I'll be posting them on my website as well
- This community is incredible at sharing and learning together...so please connect with me on Bluesky, LinkedIn, or email
- I'd love to hear about your experiences and learn from your dashboard challenges too
- And if you implement this and see results, please share them!
- [end by 16:40]
:::

@@

::: {.notes}
- [40 seconds]
- If you remember just one thing from our time together today...Dashboards don't need more features, they need fewer obstacles
- Every dropdown, every filter, should bring users closer to insight, not further from it
- [CLICK] And we can do this together...the tools exist, the knowledge exists...and it's clear from all the head nods today that you all care about creating better experiences for your users
- [CLICK] Before I close, I want to offer my heartfelt thanks to the posit::conf organizers for creating this incredible space...to the broader Shiny community for pushing the boundaries of what's possible...and most importantly, to all of you...for taking the time to be here today and for caring about your users
- Now, I'd love to hear your questions and learn about the dashboard challenges you're facing!
- [end by 17:20]
:::
